A Reproducible Data Analysis Workflow with RMarkdown, Git, Make & Docker

### 2019-09-03 abstract.Rmd
add productivity #26
minor revision

In this tutorial, we describe a principled workflow to ensure long-term reproducibility of R-based data analyses. The workflow is based on established tools and practices from software engineering. It combines the benefits of various open-source software tools including Rmarkdown, git, make and docker whose interplay ensures seamless integration of version management, (APA-conformable) dynamic report generation and full cross-platform and long-term computational reproducibility. The workflow ensures meeting the primary goals that 1) the reporting of analysis results is perfectly consistent with the actual analysis results (dynamic report generation), the analysis exactly reproduces at a later time even if the computing and software platform is updated or changed (computational reproducibility), and 3) changes at any time (during development and post-publication) are tracked, tagged, and documented while earlier versions of both data and code remain accessible. While dynamic document generation is increasingly recognized as a tool to ensure reproducibility, we demonstrate with practical examples that dynamic documents alone are not sufficient to ensure long-term computational reproducibility.

>>>>

In this tutorial, we describe a principled workflow to ensure long-term reproducibility of R-based data analyses. The workflow leverages established tools and practices from software engineering. It combines the benefits of various open-source software tools including Rmarkdown, git, make, and docker, whose interplay ensures seamless integration of version management, APA-conformable dynamic report generation and full cross-platform and long-term computational reproducibility. The workflow ensures meeting the primary goals that 1) the reporting of analysis results is perfectly consistent with the actual analysis results (dynamic report generation), the analysis exactly reproduces at a later time even if the computing and software platform is updated or changed (computational reproducibility), and 3) changes at any time (during development and post-publication) are tracked, tagged, and documented while earlier versions of both data and code remain accessible.  While the research community increasingly recognizes dynamic document generation as a tool to ensure reproducibility, we demonstrate with practical examples that dynamic documents alone are not sufficient to ensure long-term computational reproducibility. By ensuring full computational reproducibility together with version management, dependence management and literate programming,  the workflow increases scientific productivity by facilitating later reuse.

### 2019-09-03 rr-flow.Rmd
minor revision

In this tutorial, we describe a principled workflow to ensure long-term reproducibility of
R-based data analyses. In this context, reproducibility means to us that scientific results together with their computational scientific workflows are reported fully, transparently and remain sustainably available, such that the workflow can be re-run by a different person and/or later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice (such as the guidelines of the German Research Foundation; @dfg2019) and ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

@@

Sometimes external requirements make it impossible that the data can be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend that you crosscheck using checksums before loading data. To do this, a checksum must be created and filed at the time of the original analysis, and at the time of reproduction the current checksum must be compared with the stored checksum (also called a "hash", e.g., using the functions provided in package `digest`).

```{r}
x <- data.frame(VAR1=c(1,2,3,4),VAR2=c(0,4,6,9) )
checksum <- digest::sha1(x)
if (checksum != "740e88404c04af1d67ffc87b008231c093539cc1"){
  error("Wrong data file loaded or data has been tempered with!")
}
```

@@

![Source of an RMarkdown](Images/rmarkdown.png)

![Rendered result](R/rmarkdown.pdf)

@@

Even when you have obtained a version of the project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files.

@@

Since `docker`'s commands grow long and are tedius to type, we advocate to use some automatic way to generate them. Fortunatly one can use `make` to automaticly generate the `docker` commands, e.g. the (simplified) Makefile for this paper allows command after `$(run)` to be condtionally passed through docker if one calls `make DOCKER=TRUE`:

@@

build: Dockerfile
	docker build -t $(projekt) .

@@

There is a pletora of linux distributions and almost as many package managers. However to our knowledge there is no (recent) Linux where `git`, `make` and `docker` are not availible. Example for Ubuntu:

@@

Replace `apt` with your package manager. Possiblie you need elevated rights; in this case add `sudo` before that.

@@

We provide a reproducible analysis as working example via [github](https://github.com/aaronpeikert/workflow-showcase) as a testbed for users. The example shows a minimalistic analysis of the 'Considerations of Future Consequences (CFC) Scale'. The analysis demonstrates a succesful implementation of our workflow including downloading external data, comparing their state using a checksum, and running a confirmatory factor analysis on the first few items using lavaan [@lavaan2012]. Once all mentioned tools are installed on a computer, the following few steps are necessary to reproduce our analysis:

>>>>

In this tutorial, we describe a principled workflow to ensure long-term and cross-platform reproducibility of
R-based data analyses. In this context, reproducibility means to us that scientific results together with their computational scientific workflows are reported fully, transparently and remain sustainably available, such that the workflow can be re-run by a different person and/or later in time and the results will be identical to the ones intially reported [@herouxCompatibleReproducibilityTaxonomy2018; @claerboutElectronicDocumentsGive1992]. The need to ensure reproducibility directly follows from commonly accepted rules of good scientific practice (such as the guidelines of the German Research Foundation; @dfg2019) and ensuring reproducibility is a prerequisite for replicability and a means to increase trustworthiness of empirical results [@epskamp2019rep].

@@

Sometimes external requirements make it impossible that the data cannot be stored and shared together with the scripts. In most of the cases we have seen, these are either space constraints or privacy considerations. In these cases, of course, unrestricted reproducibility is not guaranteed. If this way of dividing data and scripts into different databases is unavoidable, we recommend that you crosscheck using checksums before loading data. To do this, a checksum must be created and filed at the time of the original analysis, and at the time of reproduction the current checksum must be compared with the stored checksum (also called a "hash", e.g., using the functions provided in package `digest`).

```{r}
x <- data.frame(VAR1=c(1,2,3,4),VAR2=c(0,4,6,9) )
checksum <- digest::sha1(x)
if (checksum != "740e88404c04af1d67ffc87b008231c093539cc1") 
	error("Wrong data file loaded or data has been tempered with!")
```

@@

<FIGURE 1 HERE>

@@

Even when you have obtained a version of the project, and you can ensure that this version is unchanged, you may not know how to reproduce the results, because it is unclear what has to be run in which order. This job gets comfortable with `make`, since it allows you to create (computational) recipes too (re-)create files. A downside of `make` is, that the installation on Windows is not straightforward. However, if `docker` (see later sections) is available this is not an issue, since `make` ships with (almost) every `docker` image.

@@

Since `docker`'s commands grow long and are tedius to type, we advocate to use some automatic way to generate them. Unfortunatly is the most comfortable way only availible on linux, which uses `make DOCKER=TRUE` together with a Makefile like the one used for that article.

@@

build: Dockerfile
	sudo docker build -t $(projekt) .

clean:
	Ruby/clean.rb
	rm -rf kitematic
.PHONY: clean

@@

There is a pletora of linux distribuation and almost as many package managers. However to our knowledge there is no (recent) Linux where `git`, `make` and `docker` are not availible. Example for Ubuntu:

@@

Replace `apt` with your package manager. Possiblie you need elevated rights in this case add `sudo` before that.

@@

We provide a reproducible analysis as working example via github as a testbed for users. The example shows a minimalistic analysis of the 'Considerations of Future Consequences (CFC) Scale'. The analysis demonstrates a succesful implementation of our workflow including downloading external data, comparing their state using a checksum, and running a confirmatory factor analysis on the first few items using lavaan [@lavaan2012]. Once all mentioned tools are installed on a computer, the following few steps are necessary to reproduce our analysis:
